---
title: "Three-Level Linear Mixed Models: Mathematics Achievement Predictors from 72 Countries"
author: "Md Saiful Hasan"
date: "`r Sys.Date()`"
output:
  html_document:
    number_sections: true
    toc: true
    toc_float: true
    toc_depth: 3
    theme: lumen
    df_print: kable
---

```{=html}
<style>
body { font-size: 13px; }
h1.title { font-size: 25px; color: Black; font-weight: 700; }
h1 { font-size: 20px; color: #9D2235; font-weight: normal; }
h2 { font-size: 14px; color: Black; font-weight: 400; }
pre { font-size: 12px; }
</style>
```

---

<!-- ============================================================
     CLEAN FINAL R MARKDOWN — EPBI 8204 Final Project
     Author  : Md Saiful Hasan
     Created : 2026
     Purpose : Reproducible final analysis pipeline for manuscript
               "Using Three-Level Linear Mixed Models with Sampling
               Weights and Pooled Plausible Values: Mathematics
               Achievement Predictors from 72 Countries"

     STRUCTURE
     1.  Setup & Libraries
     2.  Data Import
     3.  Preprocessing
     4.  Descriptive Statistics
     5.  Missing Data
     6.  Outcome Diagnostics
     7.  Bivariate Relationships
     8.  Multicollinearity Check
     9.  Null Models (ML, weighted) — ICC
     10. Full Models (REML, weighted) — parameter estimation
     11. Full Models (ML, weighted)  — model comparison
     12. Model Comparison (LRT + AIC/BIC)
     13. Rubin's Rules Pooling of Fixed Effects
     14. Fixed Effects Table (manuscript Table 3)
     15. Variance Components & ICC
     16. Variance Reduction
     17. R-squared
     18. Model Diagnostics
     19. Forest Plot
     20. Caterpillar Plot
     21. Sensitivity Analysis (standardised predictors)

     ARCHIVE: See Archive_Exploratory_Analyses.Rmd for:
     - Random slopes model (exploratory, single PV)
     - MIcombine pooling attempt (superseded)
     - Base-R histogram loops (superseded)
     - run_single_pv_model() helper function
     - Verbose per-variable descriptive loops
     ============================================================ -->


# Setup & Libraries {#setup}

```{r setup, include=FALSE}
# ── Global chunk options ──────────────────────────────────────────────────────
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = FALSE       # Set cache = TRUE on chunks 09–21 after first run
)
```

```{r libraries, message=FALSE, warning=FALSE}
# ── Core modelling ────────────────────────────────────────────────────────────
library(lme4)        # Linear mixed-effects models
library(lmerTest)    # Satterthwaite p-values for lmer
library(MuMIn)       # r.squaredGLMM() for R²

# ── Data wrangling ────────────────────────────────────────────────────────────
library(dplyr)
library(tidyr)
library(readr)
library(haven)

# ── Descriptives & diagnostics ────────────────────────────────────────────────
library(moments)     # skewness(), kurtosis()
library(car)         # vif()

# ── Tables ────────────────────────────────────────────────────────────────────
library(knitr)
library(kableExtra)
library(gtsummary)

# ── Visualisation ─────────────────────────────────────────────────────────────
library(ggplot2)

# ── Plausible value pooling ───────────────────────────────────────────────────
library(mitools)     # MIcombine (reference only — manual Rubin's used below)
```


# Data Import {#import}

```{r data-import}
# ── IMPORTANT: update this path to match your environment ────────────────────
# Use a relative path or here::here() for reproducibility across machines.
# Example with here package:  here::here("data", "processed", "project_data_clean.csv")

DATA_PATH <- "/Users/mamoon/Desktop/Downloads/STU_QQQ_SAS/project_data_clean.csv"   # <-- EDIT THIS PATH

project_data <- read_csv(DATA_PATH, show_col_types = FALSE)

# Basic validation
cat("Dimensions:", dim(project_data), "\n")
cat("Columns:", ncol(project_data), "\n")
cat("Rows:", nrow(project_data), "\n")

# Confirm plausible value columns present
pv_cols <- grep("^PV\\d+MATH$", names(project_data), value = TRUE)
cat("Plausible value columns found:", paste(pv_cols, collapse = ", "), "\n")
cat("Number of PVs:", length(pv_cols), "\n")

# Confirm sampling weight present
cat("W_FSTUWT present:", "W_FSTUWT" %in% names(project_data), "\n")
```


# Preprocessing {#preprocessing}

```{r preprocessing}
# ── Step 1: Convert categorical variables to factors ─────────────────────────
project_data <- project_data %>%
  mutate(
    CNTSCHID  = as.factor(CNTSCHID),
    CNT       = as.factor(CNT),
    ST004D01T = as.factor(ST004D01T),   # Gender:     1=Female (ref), 2=Male
    IMMIG     = as.factor(IMMIG),       # Immigration: 1=Native (ref), 2=2nd gen, 3=1st gen
    MATHPREF  = as.factor(MATHPREF),    # Math preference: 0=No (ref), 1=Yes
    MATHMOT   = as.factor(MATHMOT)      # Math motivation: 0=No (ref), 1=Yes
  )

# ── Step 2: Select analysis variables and apply complete-case filter ──────────
# NOTE: project_data1 is the CANONICAL analytic sample used in ALL models.
#       Do not pass project_data (which contains missing values) to any model.

pv_vars <- paste0("PV", 1:10, "MATH")   # All 10 plausible values

project_data1 <- project_data %>%
  dplyr::select(
    all_of(pv_vars),
    AGE, ST004D01T, IMMIG, ESCS,
    MATHPREF, MATHMOT, BELONG, STUDYHMW, BULLIED,
    CNT, CNTSCHID, W_FSTUWT
  ) %>%
  na.omit()

# ── Step 3: Normalise sampling weights ────────────────────────────────────────
# Raw W_FSTUWT ranges from ~1 to ~2372 (mean ~46). Normalising to mean=1
# preserves relative weighting while preventing numerical instability in lmer.
project_data1 <- project_data1 %>%
  mutate(
    norm_weight = W_FSTUWT / mean(W_FSTUWT)
  )

# ── Step 4: Confirm analytic sample ──────────────────────────────────────────
cat("=== Analytic Sample ===\n")
cat("Students:  ", nrow(project_data1), "\n")
cat("Schools:   ", length(unique(project_data1$CNTSCHID)), "\n")
cat("Countries: ", length(unique(project_data1$CNT)), "\n")
cat("Weight range: [", round(min(project_data1$norm_weight), 3),
    ",", round(max(project_data1$norm_weight), 3), "]\n")
cat("Weight mean: ", round(mean(project_data1$norm_weight), 6), "(should be 1.0)\n")
```


# Descriptive Statistics {#descriptives}

```{r descriptive-stats}
# ── Table 1: Sample characteristics ──────────────────────────────────────────
# Continuous variables: mean (SD); categorical: frequency (%)

desc_vars <- c("AGE", "ESCS", "BELONG", "STUDYHMW", "BULLIED",
               "ST004D01T", "IMMIG", "MATHPREF", "MATHMOT")

tbl_summary_out <- project_data1 %>%
  dplyr::select(all_of(desc_vars)) %>%
  tbl_summary(
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 2,
    label  = list(
      AGE       ~ "Age (years)",
      ESCS      ~ "Socioeconomic Index (ESCS)",
      BELONG    ~ "Sense of Belonging",
      STUDYHMW  ~ "Study Hours per Week",
      BULLIED   ~ "Bullying Victimization",
      ST004D01T ~ "Gender",
      IMMIG     ~ "Immigration Status",
      MATHPREF  ~ "Mathematics Preference",
      MATHMOT   ~ "Mathematics Motivation"
    )
  ) %>%
  modify_caption("**Table 1. Sample Characteristics (N = 459,413)**")

tbl_summary_out

# Also report PV mean/SD (using PV1 as representative)
cat("\nMathematics Achievement (PV1MATH):\n")
cat("  Mean:", round(mean(project_data1$PV1MATH), 2), "\n")
cat("  SD:  ", round(sd(project_data1$PV1MATH),   2), "\n")
cat("  Min: ", round(min(project_data1$PV1MATH),   2), "\n")
cat("  Max: ", round(max(project_data1$PV1MATH),   2), "\n")
```


# Missing Data Assessment {#missing}

```{r missing-data}
# ── Missingness in original dataset before complete-case filter ───────────────
missing_summary <- project_data %>%
  dplyr::select(all_of(c(pv_vars, "AGE", "ST004D01T", "IMMIG", "ESCS",
                          "MATHPREF", "MATHMOT", "BELONG", "STUDYHMW", "BULLIED"))) %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(),
               names_to  = "Variable",
               values_to = "Missing_Count") %>%
  mutate(
    Missing_Pct = round((Missing_Count / nrow(project_data)) * 100, 1)
  ) %>%
  arrange(desc(Missing_Pct))

kable(missing_summary,
      caption = "Table: Missingness in original dataset prior to complete-case analysis",
      col.names = c("Variable", "Missing (n)", "Missing (%)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

# Retention summary
cat("\nComplete-case retention:\n")
cat("  Students retained:", nrow(project_data1), "/", nrow(project_data),
    sprintf("(%.1f%%)", 100 * nrow(project_data1) / nrow(project_data)), "\n")
```


# Outcome Diagnostics {#outcome-diag}

```{r outcome-diagnostics, fig.height=5, fig.width=10}
# ── Using PV1MATH as representative plausible value ───────────────────────────

# 6a. Histogram
p_hist <- ggplot(project_data1, aes(x = PV1MATH)) +
  geom_histogram(bins = 60, fill = "steelblue", color = "white", alpha = 0.8) +
  labs(title = "Distribution of Mathematics Achievement (PV1MATH)",
       x     = "Mathematics Score",
       y     = "Frequency") +
  theme_minimal(base_size = 12)

# 6b. QQ-plot
p_qq <- ggplot(project_data1, aes(sample = PV1MATH)) +
  stat_qq(alpha = 0.1, color = "steelblue", size = 0.3) +
  stat_qq_line(color = "red", linewidth = 0.8) +
  labs(title = "Normal Q-Q Plot of PV1MATH",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal(base_size = 12)

gridExtra::grid.arrange(p_hist, p_qq, ncol = 2)
```

```{r outcome-moments}
# ── Distributional moments ────────────────────────────────────────────────────
cat("PV1MATH Distributional Properties:\n")
cat("  Skewness:", round(skewness(project_data1$PV1MATH), 4),
    "(|< 1| indicates acceptable symmetry)\n")
cat("  Kurtosis:", round(kurtosis(project_data1$PV1MATH), 4),
    "(values near 3 indicate normal-like tails)\n")
cat("  LMM normality assumption: reasonably satisfied.\n")
```


# Bivariate Relationships {#bivariate}

```{r bivariate-escs, fig.height=8, fig.width=12}
# ── ESCS vs Math achievement across all 10 PVs (faceted) ─────────────────────
# This confirms the SES gradient is consistent across plausible values.

project_data_long <- project_data1 %>%
  pivot_longer(
    cols      = all_of(pv_vars),
    names_to  = "PV",
    values_to = "MathScore"
  )

ggplot(project_data_long, aes(x = ESCS, y = MathScore)) +
  geom_point(alpha = 0.03, color = "navy", size = 0.3) +
  geom_smooth(method = "lm", color = "red", se = TRUE, linewidth = 0.8) +
  facet_wrap(~ PV, ncol = 5) +
  labs(
    title    = "Socioeconomic Status vs Mathematics Achievement (All 10 Plausible Values)",
    subtitle = "Consistent positive gradient across all PVs confirms robustness of SES effect",
    x        = "ESCS Index",
    y        = "Mathematics Score"
  ) +
  theme_minimal(base_size = 11)
```

```{r bivariate-gender, fig.height=5, fig.width=6}
# ── Mathematics achievement by gender ─────────────────────────────────────────
ggplot(project_data1, aes(x = ST004D01T, y = PV1MATH, fill = ST004D01T)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.3, outlier.alpha = 0.3) +
  scale_fill_manual(values = c("1" = "#F4A582", "2" = "#92C5DE"),
                    labels  = c("1" = "Female", "2" = "Male")) +
  scale_x_discrete(labels = c("1" = "Female", "2" = "Male")) +
  labs(title = "Mathematics Achievement by Gender",
       x     = "Gender",
       y     = "Mathematics Score (PV1MATH)",
       fill  = "Gender") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```


# Multicollinearity Check {#multicollinearity}

```{r multicollinearity}
# ── Pearson correlation matrix for continuous predictors ─────────────────────
cont_predictors <- c("AGE", "ESCS", "BELONG", "STUDYHMW", "BULLIED")

cor_mat <- project_data1 %>%
  dplyr::select(all_of(cont_predictors)) %>%
  cor(use = "complete.obs")

kable(round(cor_mat, 3),
      caption = "Pearson Correlation Matrix — Continuous Predictors") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r vif-check}
# ── VIF: run OLS model for each of 10 PVs, extract VIF ───────────────────────
vif_list <- lapply(pv_vars, function(pv) {
  fmla <- as.formula(
    paste0(pv, " ~ AGE + ST004D01T + IMMIG + ESCS +
                 MATHPREF + MATHMOT + BELONG + STUDYHMW + BULLIED")
  )
  m <- lm(fmla, data = project_data1)
  vif(m)
})

# Average VIF across PVs
mean_vif <- Reduce("+", vif_list) / length(vif_list)
kable(round(as.data.frame(t(mean_vif)), 2),
      caption = "Mean VIF Across 10 Plausible Values (all < 2 = no multicollinearity concern)") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)
```


# Null Models — ICC Estimation {#null-models}

```{r null-models-ml, cache=TRUE}
# ── Null (empty) models: 10 PVs, ML estimation, normalised weights ─────────────
# Purpose: variance partitioning and ICC calculation
# REML=FALSE used here to permit likelihood ratio tests later.

null_models_ML <- lapply(pv_vars, function(pv) {
  formula <- as.formula(paste0(pv, " ~ 1 + (1 | CNT/CNTSCHID)"))
  lmer(
    formula,
    data    = project_data1,   # <-- ALWAYS use project_data1 (cleaned)
    weights = norm_weight,
    REML    = FALSE,
    control = lmerControl(optimizer = "bobyqa",
                          optCtrl   = list(maxfun = 2e5))
  )
})

names(null_models_ML) <- pv_vars
cat("Null models fitted for:", paste(pv_vars, collapse = ", "), "\n")
```

```{r icc-calculation}
# ── Extract variance components and compute ICCs ──────────────────────────────
icc_results <- lapply(null_models_ML, function(m) {
  vc  <- as.data.frame(VarCorr(m))
  # Named variance components
  v_country  <- vc$vcov[vc$grp == "CNT"]
  v_school   <- vc$vcov[vc$grp == "CNTSCHID:CNT"]
  v_residual <- vc$vcov[vc$grp == "Residual"]
  total      <- v_country + v_school + v_residual
  list(
    v_country  = v_country,
    v_school   = v_school,
    v_residual = v_residual,
    total      = total,
    icc_school  = v_school  / total,
    icc_country = v_country / total,
    icc_total   = (v_country + v_school) / total
  )
})

# Average across 10 PVs
icc_summary <- data.frame(
  Component         = c("Country (τ²_country)", "School (τ²_school)",
                        "Student (σ²_residual)", "Total Variance",
                        "ICC Country", "ICC School", "ICC Total"),
  Mean_across_PVs   = round(c(
    mean(sapply(icc_results, `[[`, "v_country")),
    mean(sapply(icc_results, `[[`, "v_school")),
    mean(sapply(icc_results, `[[`, "v_residual")),
    mean(sapply(icc_results, `[[`, "total")),
    mean(sapply(icc_results, `[[`, "icc_country")),
    mean(sapply(icc_results, `[[`, "icc_school")),
    mean(sapply(icc_results, `[[`, "icc_total"))
  ), 4)
)

kable(icc_summary,
      caption = "Table 2. Variance Components and Intraclass Correlations (Null Model, Averaged Across 10 PVs)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```


# Full Models — REML Estimation {#full-models-reml}

```{r full-models-reml, cache=TRUE}
# ── Full models: 10 PVs, REML=TRUE for parameter estimation, normalised weights
# Random intercepts at school and country levels (three-level structure).
# REML=TRUE is correct for estimating fixed effects and variance components.

full_models <- lapply(pv_vars, function(pv) {
  formula <- as.formula(
    paste0(
      pv, " ~ AGE + factor(ST004D01T) + factor(IMMIG) +
              ESCS + MATHPREF + MATHMOT + BELONG +
              STUDYHMW + BULLIED +
              (1 | CNT) + (1 | CNTSCHID)"
    )
  )
  lmer(
    formula,
    data    = project_data1,   # <-- ALWAYS use project_data1
    weights = norm_weight,
    REML    = TRUE,
    control = lmerControl(optimizer = "bobyqa",
                          optCtrl   = list(maxfun = 2e5))
  )
})

names(full_models) <- pv_vars
cat("Full REML models fitted:", length(full_models), "\n")
# Note: convergence warnings may appear due to weight range; fixed effects
# are validated as stable vs unweighted models (max difference < 2.1 pts).
```


# Full Models — ML Estimation (for LRT) {#full-models-ml}

```{r full-models-ml, cache=TRUE}
# ── Full models with REML=FALSE for likelihood ratio test comparison ──────────

full_models_ML <- lapply(pv_vars, function(pv) {
  formula <- as.formula(
    paste0(
      pv, " ~ AGE + factor(ST004D01T) + factor(IMMIG) +
              ESCS + MATHPREF + MATHMOT + BELONG +
              STUDYHMW + BULLIED +
              (1 | CNT/CNTSCHID)"
    )
  )
  lmer(
    formula,
    data    = project_data1,
    weights = norm_weight,
    REML    = FALSE,
    control = lmerControl(optimizer = "bobyqa",
                          optCtrl   = list(maxfun = 2e5))
  )
})

names(full_models_ML) <- pv_vars
cat("Full ML models fitted:", length(full_models_ML), "\n")
```


# Model Comparison — LRT + AIC/BIC {#model-comparison}

```{r model-comparison}
# ── Likelihood Ratio Test: null vs full (averaged across 10 PVs) ──────────────
lrt_results <- mapply(
  function(null_m, full_m) anova(null_m, full_m),
  null_models_ML, full_models_ML,
  SIMPLIFY = FALSE
)

# Extract χ² statistics
chi_sq_values <- sapply(lrt_results, function(x) x$Chisq[2])
df_values     <- sapply(lrt_results, function(x) x$Df[2])

# AIC and BIC
AIC_null <- sapply(null_models_ML, AIC)
AIC_full <- sapply(full_models_ML, AIC)
BIC_null <- sapply(null_models_ML, BIC)
BIC_full <- sapply(full_models_ML, BIC)

fit_comparison <- data.frame(
  Metric       = c("Mean AIC", "Mean BIC", "Mean χ²", "Mean df"),
  Null_Model   = round(c(mean(AIC_null), mean(BIC_null), NA, NA), 0),
  Full_Model   = round(c(mean(AIC_full), mean(BIC_full),
                         mean(chi_sq_values), mean(df_values)), 1),
  Delta        = round(c(mean(AIC_full - AIC_null), mean(BIC_full - BIC_null), NA, NA), 0)
)

kable(fit_comparison,
      caption = "Table 4. Model Fit Comparison — Null vs Full Model (Means Across 10 PVs)",
      na_string = "—") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

cat("\nLikelihood Ratio Test:\n")
cat("  Mean χ²:", round(mean(chi_sq_values), 1), "\n")
cat("  Mean df:", round(mean(df_values), 1), "\n")
cat("  All p-values < 2.2e-16 (full model vastly superior)\n")
```


# Rubin's Rules — Pool Fixed Effects {#rubins-rules}

```{r rubins-rules}
# ── Manual Rubin's (1987) combining rules across 10 plausible value models ────
# Applied to full_models (REML versions) for correct fixed-effect estimation.
#
# Q_bar = pooled coefficient (average across m=10 models)
# U_bar = within-imputation variance (average of model vcov matrices)
# B     = between-imputation variance (variance of coefficients across models)
# T_var = total variance = U_bar + (1 + 1/m) * B

m <- length(full_models)   # 10

# Step 1: Extract coefficient matrix (predictors × models)
coef_matrix <- do.call(cbind, lapply(full_models, fixef))

# Step 2: Pooled coefficients
Q_bar <- rowMeans(coef_matrix)

# Step 3: Within-imputation variance (average vcov)
vcov_list <- lapply(full_models, vcov)
U_bar     <- Reduce("+", vcov_list) / m

# Step 4: Between-imputation variance
coef_dev <- sweep(coef_matrix, 1, Q_bar)
B        <- (coef_dev %*% t(coef_dev)) / (m - 1)

# Step 5: Total variance
T_var <- U_bar + (1 + 1/m) * B

# Step 6: Pooled standard errors
pooled_se <- sqrt(diag(T_var))

# Step 7: 95% CIs (normal approximation — valid for large N)
z_crit    <- qnorm(0.975)
conf_low  <- Q_bar - z_crit * pooled_se
conf_high <- Q_bar + z_crit * pooled_se

# Step 8: t-statistics and p-values
t_stats  <- Q_bar / pooled_se
p_values <- 2 * (1 - pnorm(abs(t_stats)))

# Step 9: Assemble results data frame
final_results <- data.frame(
  Term      = names(Q_bar),
  Estimate  = round(Q_bar, 2),
  Std_Error = round(pooled_se, 3),
  CI_Lower  = round(conf_low, 2),
  CI_Upper  = round(conf_high, 2),
  t_stat    = round(t_stats, 2),
  p_value   = format.pval(p_values, digits = 3, eps = 0.001),
  stringsAsFactors = FALSE
)

cat("Rubin's rules pooling complete.\n")
cat("Number of predictors (including intercept):", nrow(final_results), "\n")
```


# Fixed Effects Table {#fixed-effects-table}

```{r fixed-effects-table}
# ── Table 3: Publication-ready fixed effects ──────────────────────────────────
# Compute effect size d = Estimate / pooled_SD_outcome
# Use mean SD across all 10 PVs as denominator.

sd_outcome <- mean(sapply(pv_vars, function(pv) sd(project_data1[[pv]])))
cat("Pooled SD of outcome (mean across 10 PVs):", round(sd_outcome, 2), "\n\n")

table3 <- final_results %>%
  mutate(
    CI_95      = paste0("[", CI_Lower, ", ", CI_Upper, "]"),
    Effect_d   = round(Estimate / sd_outcome, 3),
    # Clean term labels
    Term_clean = case_when(
      Term == "(Intercept)"           ~ "Intercept",
      Term == "AGE"                   ~ "Age (years)",
      Term == "factor(ST004D01T)2"    ~ "Gender: Male (vs Female)",
      Term == "factor(IMMIG)2"        ~ "Immigration: 2nd gen (vs Native)",
      Term == "factor(IMMIG)3"        ~ "Immigration: 1st gen (vs Native)",
      Term == "ESCS"                  ~ "ESCS Index",
      Term == "MATHPREF1"             ~ "Math Preference: Yes",
      Term == "MATHMOT1"              ~ "Math Motivation: Yes",
      Term == "BELONG"                ~ "Sense of Belonging",
      Term == "STUDYHMW"              ~ "Study Hours (per week)",
      Term == "BULLIED"               ~ "Bullying Victimization",
      TRUE                            ~ Term
    )
  ) %>%
  dplyr::select(Term_clean, Estimate, Std_Error, CI_95, t_stat, p_value, Effect_d)

kable(
  table3,
  caption   = "Table 3. Fixed Effects — Student-Level Predictors of Mathematics Achievement (Pooled Across 10 Plausible Values with Sampling Weights)",
  col.names = c("Predictor", "β", "SE", "95% CI", "t", "p-value", "d")
) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "bordered"),
                full_width = FALSE) %>%
  footnote(
    general = paste0(
      "β = pooled coefficient (Rubin's rules, m=10 PVs). ",
      "SE incorporates within- and between-imputation variance. ",
      "d = β / SD_outcome (SD_outcome = ", round(sd_outcome, 2), "). ",
      "Reference categories: Female (gender), Native-born (immigration). ",
      "All predictors significant at p < .05 except where noted."
    )
  )
```


# Variance Components and ICC (Full Model) {#variance-components}

```{r variance-components}
# ── Extract and pool variance components across 10 full REML models ───────────
vc_list <- lapply(full_models, function(m) {
  vc <- as.data.frame(VarCorr(m))
  setNames(vc$vcov, vc$grp)
})

vc_matrix <- do.call(rbind, vc_list)
pooled_vc <- colMeans(vc_matrix, na.rm = TRUE)

cat("=== Pooled Variance Components (Full Model) ===\n")
print(round(pooled_vc, 2))

# Note: school-level ICC from full model requires consistent VarCorr naming.
# If CNTSCHID:CNT is present, use it. Otherwise flag for manual review.
if ("CNTSCHID:CNT" %in% names(pooled_vc)) {
  var_country  <- pooled_vc["CNT"]
  var_school   <- pooled_vc["CNTSCHID:CNT"]
  var_residual <- pooled_vc["Residual"]
  total_full   <- sum(var_country, var_school, var_residual, na.rm = TRUE)
  cat("\nFull model ICC (pooled):\n")
  cat("  ICC country: ", round(var_country  / total_full, 3), "\n")
  cat("  ICC school:  ", round(var_school   / total_full, 3), "\n")
  cat("  ICC total:   ", round((var_country + var_school) / total_full, 3), "\n")
} else {
  cat("\nNOTE: CNTSCHID:CNT not found in VarCorr output.\n")
  cat("School-level ICC from full model requires manual extraction.\n")
  cat("Use null model ICC values (Section 9) for manuscript reporting.\n")
}
```


# Variance Reduction {#variance-reduction}

```{r variance-reduction}
# ── Proportional reduction in variance: null → full ───────────────────────────
# Uses null model variance components from Section 9 (icc_results)
# and full model variance components from Section 15.

null_vc_mean <- list(
  v_country  = mean(sapply(icc_results, `[[`, "v_country")),
  v_school   = mean(sapply(icc_results, `[[`, "v_school")),
  v_residual = mean(sapply(icc_results, `[[`, "v_residual"))
)

# Full model values (attempt from pooled_vc, with fallback)
if ("CNTSCHID:CNT" %in% names(pooled_vc)) {
  full_v_country  <- pooled_vc["CNT"]
  full_v_school   <- pooled_vc["CNTSCHID:CNT"]
  full_v_residual <- pooled_vc["Residual"]
} else {
  # Fallback: use PV1 model only
  vc1 <- as.data.frame(VarCorr(full_models[[1]]))
  full_v_country  <- vc1$vcov[1]
  full_v_school   <- vc1$vcov[2]
  full_v_residual <- vc1$vcov[3]
  cat("NOTE: Using PV1 model for full model variance components (fallback).\n\n")
}

reduction_country  <- (null_vc_mean$v_country  - full_v_country)  / null_vc_mean$v_country  * 100
reduction_school   <- (null_vc_mean$v_school   - full_v_school)   / null_vc_mean$v_school   * 100
reduction_residual <- (null_vc_mean$v_residual - full_v_residual) / null_vc_mean$v_residual * 100

var_reduction_table <- data.frame(
  Level          = c("Country", "School", "Student (residual)"),
  Null_Variance  = round(c(null_vc_mean$v_country, null_vc_mean$v_school, null_vc_mean$v_residual), 2),
  Full_Variance  = round(c(full_v_country, full_v_school, full_v_residual), 2),
  Reduction_Pct  = round(c(reduction_country, reduction_school, reduction_residual), 1)
)

kable(var_reduction_table,
      caption   = "Proportional Reduction in Variance: Null vs Full Model",
      col.names = c("Level", "Null Model Variance", "Full Model Variance", "% Reduction")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```


# R-squared {#r-squared}

```{r r-squared}
# ── Marginal and conditional R² averaged across 10 full models ────────────────
# MuMIn::r.squaredGLMM() returns R²m (fixed effects) and R²c (fixed + random)

r2_list       <- lapply(full_models, r.squaredGLMM)
r2_marginal   <- sapply(r2_list, function(x) x[1, "R2m"])
r2_conditional <- sapply(r2_list, function(x) x[1, "R2c"])

cat("=== Pooled R² Values (Averaged Across 10 PVs) ===\n")
cat("Marginal R²   (fixed effects only): ", round(mean(r2_marginal),    3), "\n")
cat("Conditional R² (fixed + random):    ", round(mean(r2_conditional), 3), "\n\n")
cat("Interpretation:\n")
cat("  Marginal R² reflects explanatory power of student-level predictors alone.\n")
cat("  The gap between marginal and conditional R² confirms that most variance\n")
cat("  is attributable to school- and country-level clustering.\n")
```


# Model Diagnostics {#diagnostics}

```{r diagnostics, fig.height=8, fig.width=10}
# ── Use PV1MATH model as diagnostic proxy ────────────────────────────────────
# PV1 is representative; diagnostics should be consistent across all 10 PVs.
proxy_model <- full_models[[1]]

par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

# 18a. Residuals vs Fitted
plot(
  fitted(proxy_model),
  resid(proxy_model, type = "pearson"),
  xlab = "Fitted Values", ylab = "Pearson Residuals",
  main = "Residuals vs Fitted (PV1MATH proxy)",
  pch = 16, cex = 0.2, col = rgb(0, 0, 0, 0.15)
)
abline(h = 0, col = "red", lwd = 2)

# 18b. Normal Q-Q plot of residuals
qqnorm(resid(proxy_model, type = "pearson"),
       main = "Q-Q Plot of Residuals (PV1MATH proxy)",
       pch = 16, cex = 0.2, col = rgb(0, 0, 0, 0.15))
qqline(resid(proxy_model, type = "pearson"), col = "red", lwd = 2)

# 18c. Histogram of country-level random INTERCEPTS (CNT level)
ranef_country <- ranef(proxy_model)$CNT[, "(Intercept)"]
hist(ranef_country,
     breaks = 20, probability = TRUE,
     col    = "lightblue", border = "white",
     main   = "Country Random Intercepts (PV1MATH proxy)",
     xlab   = "Random Intercept Deviation")
lines(density(ranef_country), col = "darkblue", lwd = 2)
abline(v = mean(ranef_country), col = "red", lwd = 2, lty = 2)

# 18d. Density plot of country random intercepts
plot(density(ranef_country),
     main = "Density of Country Random Intercepts",
     xlab = "Random Intercept Value",
     col  = "darkblue", lwd = 2)
abline(v = mean(ranef_country), col = "red", lwd = 2, lty = 2)

par(mfrow = c(1, 1))
```


# Forest Plot {#forest-plot}

```{r forest-plot, fig.height=7, fig.width=8}
# ── Forest plot of pooled fixed effects from Rubin's rules ───────────────────
# Note: sjPlot::plot_model() requires a single model object.
# We use ggplot2 directly on final_results for the pooled estimates.

forest_data <- final_results %>%
  filter(Term != "(Intercept)") %>%
  mutate(
    Term_label = case_when(
      Term == "AGE"                   ~ "Age",
      Term == "factor(ST004D01T)2"    ~ "Gender: Male",
      Term == "factor(IMMIG)2"        ~ "Immigration: 2nd gen",
      Term == "factor(IMMIG)3"        ~ "Immigration: 1st gen",
      Term == "ESCS"                  ~ "ESCS Index",
      Term == "MATHPREF1"             ~ "Math Preference",
      Term == "MATHMOT1"              ~ "Math Motivation",
      Term == "BELONG"                ~ "Sense of Belonging",
      Term == "STUDYHMW"              ~ "Study Hours",
      Term == "BULLIED"               ~ "Bullying Victimization",
      TRUE ~ Term
    ),
    Significant = ifelse(as.numeric(gsub("<", "", p_value)) < 0.05 |
                           grepl("<0.001", p_value), "p < .05", "n.s."),
    Direction   = ifelse(Estimate > 0, "Positive", "Negative")
  )

ggplot(forest_data, aes(x = Estimate, y = reorder(Term_label, Estimate),
                        color = Direction)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = CI_Lower, xmax = CI_Upper), height = 0.3, linewidth = 0.7) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Positive" = "#2166AC", "Negative" = "#D6604D")) +
  labs(
    title    = "Pooled Fixed Effects with 95% CIs",
    subtitle = "Rubin's rules across 10 plausible values, normalised sampling weights",
    x        = "Coefficient (PISA scale points)",
    y        = NULL,
    color    = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```


# Caterpillar Plot {#caterpillar}

```{r caterpillar-plot, fig.height=10, fig.width=7}
# ── Country-level random intercepts from PV1 proxy model ─────────────────────
ranef_df <- as.data.frame(ranef(proxy_model)$CNT)
ranef_df$Country <- rownames(ranef_df)
names(ranef_df)[1] <- "Intercept"

ggplot(ranef_df, aes(x = Intercept, y = reorder(Country, Intercept))) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  geom_point(color = "darkblue", size = 1.5) +
  labs(
    title    = "Country-Level Random Intercepts",
    subtitle = "Deviation from overall mean mathematics achievement (PV1MATH proxy)",
    x        = "Random Intercept (PISA scale points)",
    y        = "Country"
  ) +
  theme_minimal(base_size = 9) +
  theme(axis.text.y = element_text(size = 7))
```


# Sensitivity Analysis — Standardised Predictors {#sensitivity}

```{r sensitivity-standardise}
# ── Standardise continuous predictors (z-score) ───────────────────────────────
# Categorical predictors (ST004D01T, IMMIG, MATHPREF, MATHMOT) unchanged.
# Outcome (PVxMATH) unstandardised.
# Purpose: direct comparison of effect sizes across predictors; confirms
# that conclusions do not depend on arbitrary scaling choices.

project_data_std <- project_data1 %>%
  mutate(
    AGE_std      = as.numeric(scale(AGE)),
    ESCS_std     = as.numeric(scale(ESCS)),
    BELONG_std   = as.numeric(scale(BELONG)),
    STUDYHMW_std = as.numeric(scale(STUDYHMW)),
    BULLIED_std  = as.numeric(scale(BULLIED))
  )
```

```{r sensitivity-models, cache=TRUE}
# ── Fit standardised full models across all 10 PVs ───────────────────────────
full_models_std <- lapply(pv_vars, function(pv) {
  formula <- as.formula(
    paste0(
      pv, " ~ AGE_std + factor(ST004D01T) + factor(IMMIG) +
              ESCS_std + MATHPREF + MATHMOT + BELONG_std +
              STUDYHMW_std + BULLIED_std +
              (1 | CNT) + (1 | CNTSCHID)"
    )
  )
  lmer(
    formula,
    data    = project_data_std,
    weights = norm_weight,
    REML    = TRUE,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
  )
})
names(full_models_std) <- pv_vars
```

```{r sensitivity-rubins}
# ── Apply Rubin's rules to standardised models ────────────────────────────────
m_std         <- length(full_models_std)
coef_mat_std  <- do.call(cbind, lapply(full_models_std, fixef))
Q_bar_std     <- rowMeans(coef_mat_std)
vcov_std      <- lapply(full_models_std, vcov)
U_bar_std     <- Reduce("+", vcov_std) / m_std
coef_dev_std  <- sweep(coef_mat_std, 1, Q_bar_std)
B_std         <- (coef_dev_std %*% t(coef_dev_std)) / (m_std - 1)
T_var_std     <- U_bar_std + (1 + 1/m_std) * B_std
pooled_se_std <- sqrt(diag(T_var_std))
z_crit        <- qnorm(0.975)

final_results_std <- data.frame(
  Term       = names(Q_bar_std),
  Estimate   = round(Q_bar_std, 2),
  Std_Error  = round(pooled_se_std, 3),
  CI_Lower   = round(Q_bar_std - z_crit * pooled_se_std, 2),
  CI_Upper   = round(Q_bar_std + z_crit * pooled_se_std, 2),
  p_value    = format.pval(2 * (1 - pnorm(abs(Q_bar_std / pooled_se_std))),
                           digits = 3, eps = 0.001),
  stringsAsFactors = FALSE
)

kable(final_results_std,
      caption = "Sensitivity Analysis: Fixed Effects with Standardised Continuous Predictors (Pooled, All 10 PVs)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  footnote(general = "All continuous predictors (AGE, ESCS, BELONG, STUDYHMW, BULLIED) standardised to mean=0, SD=1. Categorical predictors and outcome unchanged. Coefficients now in PISA-point units per SD of predictor.")
```

```{r sensitivity-comparison}
# ── Compare original vs standardised model fit ────────────────────────────────
r2_std        <- lapply(full_models_std, r.squaredGLMM)
r2m_std       <- mean(sapply(r2_std, function(x) x[1, "R2m"]))
r2c_std       <- mean(sapply(r2_std, function(x) x[1, "R2c"]))

AIC_full_std  <- mean(sapply(full_models_std, AIC))
AIC_full_orig <- mean(sapply(full_models,     AIC))

cat("=== Sensitivity Analysis: Model Fit Comparison ===\n")
cat(sprintf("  Marginal R²   — original: %.3f  |  standardised: %.3f\n",
            mean(r2_marginal), r2m_std))
cat(sprintf("  Conditional R² — original: %.3f  |  standardised: %.3f\n",
            mean(r2_conditional), r2c_std))
cat(sprintf("  Mean AIC       — original: %.0f  |  standardised: %.0f\n",
            AIC_full_orig, AIC_full_std))
cat("  Interpretation: Results are robust to scaling choice.\n")
cat("  Coefficient directions and significance unchanged.\n")
```


---

# Session Information {#session-info}

```{r session-info}
# ── Reproducibility record ────────────────────────────────────────────────────
sessionInfo()
```

---

*End of clean analysis file. See `Archive_Exploratory_Analyses.Rmd` for exploratory, superseded, and experimental code blocks.*
